Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-07-14 ms/manuscript.Rmd
Update MS Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(tidyverse)
library(drake)
library(patchwork)
library(kableExtra)
library(here)
library(magrittr)
library(scales)
```

@@

We selected several traits relevant to epiphytic growth (Table \@ref(tab:traits-used) and Figure \@ref(fig:traits-used-image)). We preferred morphological traits over physiological traits for our study because data for morphological traits are easier to gather at the scale of our study (\circa \SI{130}{\species} distributed over an elevational gradient from \SIrange{200}{2000}{\meter}). In addition, plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010]. Studying the relationships between ecology and morphological traits is useful from a paleobotanical and climate change perspective, as trends in morphological trait variation in the fossil record can be used to gain insight into past climatic conditions [@Chaloner1997; @McElwain2007a; @McElwain2007b; @Royer2008].

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; ref table 1). Trait states were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017], grown from spores of known species in the lab, or taken from the literature. For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate. Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (ref traits figure). Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea ($\n = 130$ \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, and 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. The probability of obtaining the observed value of \D under either model is then calculated.

@@

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. The first model tests the null hypothesis that rates of evolution of the two traits are independent. The second model tests the alternative hypothesis that the rates of evolution of one trait depends on the other. The likelihoods of obtaining the observed data are calculated for each model, then compared using a log likelihood test. A significantly better fit of the dependent model indicates that evolution of the two traits is correlated. To verify the effect of analysis method on our results, we performed alternative tests for correlation of growth habit with functional traits using phylogenetic generalized linear mixed models (PGLMMs; see Methods S1 and Tables S1 and S2).

@@

@@

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to plant size, including stipe and frond length, frond width, and rhizome diameter. Stipe length in particular differed strongly between epiphytes and terrestrial species (Figure \@ref(fig:cwm-div), TABLE PIC SPOROS). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; Creese2011]. However, given that many of these size-related traits scale together in ferns [@Arcand2008; Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

@@

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (Figure \@ref(fig:cwm-div), TABLE PIC SPOROS). SLA is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a]. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, *Davallia anderssonii* Mett. ex Kuhn, \SI{`r smallest_sla_1`} {\square\meter\per\kilo\gram}; *Ctenopterella blechnoides* (Grev.) Parris, \SI{`r smallest_sla_2`} {\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (Figure \@ref(fig:cwm-div), TABLE PIC SPOROS). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies [@Petter2016]. For example, @Zhu2016 did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of @Zhu2016. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (Figure \@ref(fig:cwm-div)).

@@

Only gametophyte morphotype was found to have a significant association with epiphytic growth. Morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived ($< \SI{1}{\year}$), cordate gametophytes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate (\eg, elongate or filamentous) gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; Farrar2008]. The complex, three-dimensional structure of non-cordate gametophytes slows their drying rates, and may be an adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013], and promote out-crossing [@Farrar2008]. Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant while controlling for phylogeny. @Sundue2015 found that non-cordate morphology is significantly correlated with increased diversification rates within Polypodiaceae, consistent with the hypothesis that this trait promoted diversification within the epiphytic niche.

@@

Presence of hairs and glands has strong phylogenetic signal (REF TABLE phylosig, REF TABLE supp pglmm gameto), and these traits are correlated with certain taxonomic groups [@Stokey1951; @Nayar1971]. Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], we did not detect a correlation between hair production and epiphytic growth. Similarly, production of glands does not appear to be correlated with epiphytic growth. Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established [@Johnson2012]. Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes.

@@

Results for functional trait richness (FRic) supported our hypotheses: we observed lower FRic in epiphytic relative to terrestrial communities overall, and a significant positive correlation of FRic with humidity in epiphytes (\ie, decreased FRic at drier, low elevation plots and increased FRic at more humid, high elevation plots; Figure \@ref(fig:comm-div)). However, results for the other two measures of functional diversity, evenness (FEve) and divergence (FDiv) did not match our predictions. Rather, FEve decreased with elevation (but not minimum RH) in terrestrial communities, and FDiv was lower in terrestrial communities overall while showing no trend along the gradient  (Figure \@ref(fig:comm-div)). The lack of a clear pattern in functional evenness and divergence suggests that within each trait pool, there is no significant difference in trait spacing between epiphytes and terrestrial species. This could be maintained by niche partitioning within epiphytic and terrestrial habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented [@Zotz2007]. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, @WatkinsJr.2006a found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both, and no species that overlapped between canopy and terrestrial habitats. It is possible that epiphytic growth per se limits overall morphological variation in epiphytic ferns, but that species with different traits are then sorted into different niches.

In addition to their decreased functional diversity, we also found that epiphytic communities are more phylogenetically clustered relative to terrestrial communities overall (Figure \@ref(fig:comm-div)). However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient. It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends. When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, @Nitta2017 detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation. @Kluge2011 also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group. They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

@@

Figure \@ref(fig:traits-used-image): Examples of traits used in this study. (**a**) Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinna pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. (**b**) Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars 1 mm except for glands, which is 0.1 mm. Photographs by J. H. Nitta.

Figure \@ref(fig:climate): Selected microclimatic variables of study plots along an elevational gradient from \SIrange{200}{1200}\meter on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from July 7, 2013 to July 5, 2014, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at ca. 2 m on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval \textless \num{.05}

Figure \@ref(fig:pca): Principal components analysis (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. **(a, b)** PC loadings. **(c, d)** species scores. **(a)** and **(c)** are standard PCA, **(b)** and **(d)** are phylogenetic PCA. Epiphytes in green, terrestrial species in brown. Quantitative traits only, including only species with no missing observations for any traits ($\n = 104$ \si{\species}).

Figure \@ref(fig:tree): Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Relative value of quantitative (sporophyte) traits in heat gradient: low values are lighter, high values are darker. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Trends for three leaf size traits (stipe length, frond length, and frond with) were quantitatively very similar, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see Table \@ref(tab:traits-used). Major Cretaceous epiphytic radiations \sensu @Schuettpelz2009 labeled on tree in green: H, hymenophylloids; T, trichomanoids, V, vittarioids; A, asplenioids; E, elaphoglossoids; P, polygrammoids (trichomanoids diversified prior to the Cretaceous but are included as an important extant epiphytic clade).

Figure \@ref(fig:cwm-div): Change in selected functional traits of ferns on Moorea, French Polynesia across environmental gradients and between growth habit types (epiphytic or terrestrial). Epiphytic values in green, terrestrial values in brown. Scatterplots show community-weighted mean (CWM) values for each trait plotted against either minimum relative humidity ($\n = 13$ communities each for epiphytes and terrestrial species) or elevation ($\n = 17$ communities each). Trendlines indicate significant relationships determined by a linear model at $\pval<0.05$. Boxplots show median values (bold lines) by growth habit (epiphytic or terrestrial) across all communities (epiphytic communities, $\n = 17$; terrestrial communities, $\n = 17$). Lower and upper hinges correspond to first and third quartiles, and whiskers extend to values within 1.5 $\times$ the interquartile range. Asterisks indicate statistical significance of two-sided \tval-test for differences between epiphytic and terrestrial species; $*** = \pval<0.001$; $** = \pval<0.01$; $* = \pval<0.05$. Trends for three leaf size traits (stipe length, frond length, and frond with) were quantitatively very similar, so of these we only present stipe length.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE)
library(drake)
library(patchwork)
library(knitr)
library(kableExtra)
library(here)
library(magrittr)
library(scales)
library(tidyverse)
library(conflicted)

conflict_prefer("expand", "tidyr")
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

# Function for rounding with trailing zeros
round_t <- function (x, digits) round(x, digits) %>% sprintf(glue::glue("%.{digits}f"), .)

# Function for formatting p-values: round to 3 digits,
# or just say "< 0.001"
format_pval <- function (x) {
  case_when(
    x < 0.001 ~ "< 0.001",
    TRUE ~ round(x, 3) %>% as.character()
  )
}

@@

We selected several traits relevant to epiphytic growth (\tab \@ref(tab:traits-used) and \fig \@ref(fig:traits-used-image)). We preferred morphological traits over physiological traits for our study because data for morphological traits are easier to gather at the scale of our study (\circa \SI{130}{\species} distributed over an elevational gradient from \SIrange{200}{2000}{\meter}). In addition, plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010]. Studying the relationships between ecology and morphological traits is useful from a paleobotanical and climate change perspective, as trends in morphological trait variation in the fossil record can be used to gain insight into past climatic conditions [@Chaloner1997; @McElwain2007a; @McElwain2007b; @Royer2008].

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; \tab \@ref(tab:traits-used)). Trait states were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017], grown from spores of known species in the lab, or taken from the literature. For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate. Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (\fig \@ref(fig:traits-used-image)). Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea ($\n = 130$ \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. The probability of obtaining the observed value of \D under either model is then calculated.

@@

We detected significant phylogenetic signal in most traits, so we tested for differences between traits of epiphytic vs. terrestrial species using methods that can test for significant differences between traits associated with a binary character while accounting for phylogeny. For quantitative traits, we used the \function{brunch} function in the \package{caper} package [@Orme2013], which calculates phylogenetically independent contrasts in a quantitative character between alternative states of a binary trait. For qualitative (binary) traits, we used Pagel's test of correlated evolution [@Pagel1994] as implemented with the \function{fitPagel} function in the \package{phytools} package [@Revell2012]. This method compares the likelihood between two alternative models for a pair of binary traits. The first model tests the null hypothesis that rates of evolution of the two traits are independent. The second model tests the alternative hypothesis that the rates of evolution of one trait depends on the other. The likelihoods of obtaining the observed data are calculated for each model, then compared using a log likelihood test. A significantly better fit of the dependent model indicates that evolution of the two traits is correlated. To verify the effect of analysis method on our results, we performed alternative tests for correlation of growth habit with functional traits using phylogenetic generalized linear mixed models (PGLMMs; see Methods S1 and \tabs S1 and S2).

@@

### Principal Components Analysis of Trait Data

```{r pca-numbers}
# Extract summary stats for mention in text
loadd(pca_results)

# Cumulative amount of variance explained by first two axes, standard PCA
pca_std_cum_var <-
  pca_results$variance %>% 
  filter(analysis_type == "standard", variance_type == "Cumulative Proportion") %>%
  pull(PC2) %>%
  signif(2)

# Cumulative amount of variance explained by first two axes, phylogenetic PCA
pca_phy_cum_var <-
  pca_results$variance %>% 
  filter(analysis_type == "phylogenetic", variance_type == "Cumulative Proportion") %>%
  pull(PC2) %>%
  signif(2)
```

<!-- CHECK PCA CODE: missing dissection -->

Terrestrial and epiphytic species are partly differentiated in trait space, occupying mostly distinct areas but overlapping in the middle regardless of method used (standard or phylogenetic PCA) (\fig \@ref(fig:pca)(c, d)). The first two PC axes explained `r percent(pca_std_cum_var)` and `r percent(pca_phy_cum_var)` of the variation in the trait data when analyzed using standard and phylogenetic PCA, respectively. The trait loadings were similar between the two analyses, except that most switched axes; for example, SLA was loaded on PC1 in the standard analysis but PC2 in the phylogenetic analysis (REF SUPP TABLE pca).

### Phylogenetic Signal

```{r physig-numbers}
loadd(phylosig_binary_traits)
habit_d <- phylosig_binary_traits %>% filter(trait == "habit") %>% pull(D) %>% round_t(2)
hairs_d <- phylosig_binary_traits %>% filter(trait == "hairs") %>% pull(D) %>% round_t(2)
gemmae_d <- phylosig_binary_traits %>% filter(trait == "gemmae") %>% pull(D) %>% round_t(2)
glands_d <- phylosig_binary_traits %>% filter(trait == "glands") %>% pull(D) %>% round_t(2)
morphotype_d <- phylosig_binary_traits %>% filter(trait == "morphotype") %>% pull(D) %>% round_t(2)
```

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), \fig \@ref(fig:tree)), and for quantitative traits, different results were obtained for $\lambda$ and K. When measured with $\lambda$, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of $\lambda$ close to 1; only number of pinna pairs had $\lambda$ close to zero (\tab \@ref(tab:phylosig-cont)). However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of $\K < 1$ (\tab \@ref(tab:phylosig-cont)). All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmmae \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under a scenario of evolution by BM (\tab \@ref(tab:phylosig-binary)). Phylogenetic generalized linear mixed models (PGLMMs) also indicated significant signal in all gametophyte traits (REF SUPP TABLE). Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)).

### Correlation of Traits with Growth Habit

```{r corr-numbers}
loadd(pic_results)
stipe_pval <- pic_results %>% filter(trait == "stipe") %>% pull(p.value) %>% signif(2)
width_pval <- pic_results %>% filter(trait == "width") %>% pull(p.value) %>% signif(2)
dissection_pval <- pic_results %>% filter(trait == "dissection") %>% pull(p.value) %>% signif(2)

loadd(correlated_evo_test)
morphotype_pval <- correlated_evo_test %>% filter(trait == "morphotype") %>% pull(pval) %>% signif(2)

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \@ref(tab:pics)). Epiphytic species tend to have smaller values for all traits (\tab \@ref(tab:pics); contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval = `r stipe_pval`), as well as frond width (\pval = `r width_pval`) and degree of laminae dissection (\pval = `r dissection_pval`). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval = `r morphotype_pval`; \tab \@ref(tab:corr-evo)). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (REF SUPP TABLE).

@@

The sporophyte traits that differed most strongly between epiphytic and terrestrial taxa were related to plant size, including stipe and frond length, frond width, and rhizome diameter. Stipe length in particular differed strongly between epiphytes and terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). Stipe length may be a critical trait controlling whole leaf hydraulic conductance; since ferns have resistive stipes in general, species growing in water-limited habitats are expected to decrease stipe length to minimize resistance [@WatkinsJr.2010]. Our results support previous non-phylogenetically controlled studies that found a prevalence of short stipes in epiphytic ferns [@WatkinsJr.2010; Creese2011]. However, given that many of these size-related traits scale together in ferns [@Arcand2008; Creese2011], careful experimental manipulations are needed to distinguish the role (or lack thereof) of each in epiphytic growth.

@@

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics). Specific leaf area is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a]. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, *Davallia anderssonii* Mett. ex Kuhn, \SI{`r smallest_sla_1`} {\square\meter\per\kilo\gram}; *Ctenopterella blechnoides* (Grev.) Parris, \SI{`r smallest_sla_2`} {\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (\fig \@ref(fig:cwm-div), \tab \@ref(tab:pics)). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies [@Petter2016]. For example, @Zhu2016 did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of @Zhu2016. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig \@ref(fig:cwm-div)).

@@

Only gametophyte morphotype was found to have a significant association with epiphytic growth. Morphotype was previously observed to correlate with life history and habitat in ferns: terrestrial species tend to have short-lived ($< \SI{1}{\year}$), cordate gametophytes that establish following disturbance and rapidly produce sporophytes, whereas epiphytic species tend to have non-cordate (\eg, elongate or filamentous) gametophytes that persist over multiple growing seasons and produce sporophytes more slowly [@WatkinsJr.2007a; @Farrar2008]. The complex, three-dimensional structure of non-cordate gametophytes slows their drying rates, and may be an adaptation to drought-prone epiphytic habitats [@WatkinsJr.2007c; @Pittermann2013], and promote out-crossing [@Farrar2008]. Our study is the first to our knowledge to demonstrate that the correlation between non-cordate morphology and epiphytism in ferns is significant while controlling for phylogeny. @Sundue2015 found that non-cordate morphology is significantly correlated with increased diversification rates within Polypodiaceae, consistent with the hypothesis that this trait promoted diversification within the epiphytic niche.

@@

Presence of hairs and glands has strong phylogenetic signal (\tabs \@ref(tab:phylosig-binary), REF TABLE supp pglmm gameto), and these traits are correlated with certain taxonomic groups [@Stokey1951; @Nayar1971]. Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], we did not detect a correlation between hair production and epiphytic growth. Similarly, production of glands does not appear to be correlated with epiphytic growth. Wax-secreting glands are prevalent in both sporophytes and gametophytes of many notholaenid ferns, a xeric-adapted clade, but their functional significance has not been established [@Johnson2012]. Further tests are needed to determine if such glands serve a functional role in preventing water loss in desert ferns and epiphytes.

@@

Results for functional trait richness (FRic) supported our hypotheses: we observed lower FRic in epiphytic relative to terrestrial communities overall, and a significant positive correlation of FRic with humidity in epiphytes (\ie, decreased FRic at drier, low elevation plots and increased FRic at more humid, high elevation plots; \fig \@ref(fig:comm-div)). However, results for the other two measures of functional diversity, evenness (FEve) and divergence (FDiv) did not match our predictions. Rather, FEve decreased with elevation (but not minimum RH) in terrestrial communities, and FDiv was lower in terrestrial communities overall while showing no trend along the gradient  (\fig \@ref(fig:comm-div)). The lack of a clear pattern in functional evenness and divergence suggests that within each trait pool, there is no significant difference in trait spacing between epiphytes and terrestrial species. This could be maintained by niche partitioning within epiphytic and terrestrial habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk [@Hietz1995]. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented [@Zotz2007]. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, @WatkinsJr.2006a found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both, and no species that overlapped between canopy and terrestrial habitats. It is possible that epiphytic growth per se limits overall morphological variation in epiphytic ferns, but that species with different traits are then sorted into different niches.

In addition to their decreased functional diversity, we also found that epiphytic communities are more phylogenetically clustered relative to terrestrial communities overall (\fig \@ref(fig:comm-div)). However, we did not detect significant trends in phylogenetic community structure of epiphytic and terrestrial communities along the elevational gradient. It is possible that the scale of the current study (from \SIrange{200}{1200}{\meter}) prevented the detection of such trends. When the plots in this study were analyzed together with additional plots from Tahiti reaching a maximum altitude of just over \SI{2000}{\meter}, @Nitta2017 detected increasing phylogenetic clustering in fern sporophyte communities (epiphytes and terrestrial species together) with elevation. @Kluge2011 also compared phylogenetic and trait diversity between epiphytic and terrestrial ferns over an elevational gradient from \SIrange{100}{3400}{\meter} in Costa Rica, and found trait overdispersion at mid-elevations for epiphytes but no trends in phylogenetic diversity for either group. They accounted this to evolutionary plasticity in traits, but did not quantify the degree of phylogenetic signal in the traits they studied.

@@

**\fig \@ref(fig:traits-used-image)**: Examples of traits used in this study. (**a**) Sporophyte traits. All sporophyte traits were quantitative, including frond length and width, stipe length, rhizome dia., number of pinna pairs (in this example, nine) and degree of lamina dissection (in this example, once-pinnate). Specific leaf area (ratio of area to mass of leaf lamina) not depicted. (**b**) Gametophyte traits. Morphotype was defined as a binary trait, either cordate or non-cordate. Examples of non-cordate morphotypes include ribbon (thallus elongate and two-dimensional) and filamentous (thallus single lines of cells). Other gametophyte traits (gemmae, hairs, and glands) were scored as present or absent. Arrows point out instances of each binary trait. Scalebars 1 mm except for glands, which is 0.1 mm. Photographs by J. H. Nitta.

**\fig \@ref(fig:climate)**: Selected microclimatic variables of study plots along an elevational gradient from \SIrange{200}{1200}\meter  on Moorea, French Polynesia. Relative humidity (RH) and temperature were recorded every \SI{15}{\minute} with dataloggers from July 7, 2013 to July 5, 2014, and overall means calculated for daily mean, minimum, maximum, and standard deviation (SD). 'Epiphytic' dataloggers were placed at ca. 2 m on trees, and 'terrestrial' dataloggers placed at ground level. Color indicates growth habit: epiphytic dataloggers in green, terrestrial dataloggers in brown. Trendlines fitted using linear models with an interaction between growth habit and elevation (see Methods); all trendlines significant at \pval \textless \num{.05}

**\fig \@ref(fig:pca)**: Principal components analysis (PCA) of traits related to epiphytic growth in ferns from Moorea, French Polynesia. **(a, b)** PC loadings. **(c, d)** species scores. **(a)** and **(c)** are standard PCA, **(b)** and **(d)** are phylogenetic PCA. Epiphytes in green, terrestrial species in brown. Quantitative traits only, including only species with no missing observations for any traits ($\n = 104$ \si{\species}).

**\fig \@ref(fig:tree)**: Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Relative value of quantitative (sporophyte) traits in heat gradient: low values are lighter, high values are darker. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Trends for three leaf size traits (stipe length, frond length, and frond with) were quantitatively very similar, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see \tab \@ref(tab:traits-used). Major Cretaceous epiphytic radiations \sensu @Schuettpelz2009 labeled on tree in green: H, hymenophylloids; T, trichomanoids, V, vittarioids; A, asplenioids; E, elaphoglossoids; P, polygrammoids (trichomanoids diversified prior to the Cretaceous but are included as an important extant epiphytic clade).

**\fig \@ref(fig:cwm-div)**: Change in selected functional traits of ferns on Moorea, French Polynesia across environmental gradients and between growth habit types (epiphytic or terrestrial). Epiphytic values in green, terrestrial values in brown. Scatterplots show community-weighted mean (CWM) values for each trait plotted against either minimum relative humidity ($\n = 13$ communities each for epiphytes and terrestrial species) or elevation ($\n = 17$ communities each). Trendlines indicate significant relationships determined by a linear model at $\pval<0.05$. Boxplots show median values (bold lines) by growth habit (epiphytic or terrestrial) across all communities (epiphytic communities, $\n = 17$; terrestrial communities, $\n = 17$). Lower and upper hinges correspond to first and third quartiles, and whiskers extend to values within 1.5 $\times$ the interquartile range. Asterisks indicate statistical significance of two-sided \tval-test for differences between epiphytic and terrestrial species; $*** = \pval<0.001$; $** = \pval<0.01$; $* = \pval<0.05$. Trends for three leaf size traits (stipe length, frond length, and frond with) were quantitatively very similar, so of these we only present stipe length.
